Bow shocks, bow waves, and dust waves

### 2018-05-27 papers/Papers1234/sec-dust-bow-intro.tex
Add new figures for the gas-grain drag

@@

The drag force on a charged dust grain moving at a relative speed
\(w\) through a plasma has contributions from both direct collisions
and from electrostatic Coulomb interactions with ions and electrons.
We use the expressions in \citet{Draine:1979a}, equations~(4)--(6),
summing over protons, electrons, and helium ions.\footnote{Helium is
  assumed to be 50\% neutral and 50\% singly ionized, leading to only
  a small contribution to the drag force.  For much hotter stars, such
  as the central stars of planetary nebulae, helium may be doubly
  ionized, which leads to a more significant drag contribution,
  especially around \(w \approx \SI{5}{km.s^{-1}}\).} %
Results are shown in Figure~\ref{fig:gas-grain-drag-photoionized} for
different values of the dimensionless grain potential:
\(\phi = e^2 Z\grain / a kT\), which is the electrostatic potential
energy of a unit charge at the surface of a grain of charge
\(Z\grain\) and radius \(a\), in units of the characteristic thermal
energy of a gas particle.  The drag force \(f\drag\) is put in dimensionless units by dividing by
a characteristic force:
\begin{equation}
  \label{eq:fstar}
  f_* = 2 n k T \cdot \pi a^2 \ , 
\end{equation}
which is approximately\footnote{%
  To simplify the exposition, the gas pressure in this section is
  calculated assuming a fully ionized, pure hydrogen plasma, yielding
  \(P\gas = 2 n k T\). For typical ISM abundances, the contribution of
  helium and its corresponding electrons gives a correction to this of
  order 5\%.  The required modifications when the stream is neutral
  are discussed later.  } %
the ionized gas pressure multiplied by the grain geometric cross
section.

For grains with low electric charge, \(\phi^2 \ll 1\), the drag force is
dominated by direct collisions of protons with the grain.  The gas
collisional mean free path is much larger than the grain size, so the
drag is in the Epstein regime \citep{Weidenschilling:1977b}.  This is
illustrated by the \(\phi = 0.25\) case (blue line) in
Figure~\ref{fig:gas-grain-drag-photoionized}. As the relative
gas--grain slip speed, \(w\), increases, \(f\drag\) first increases
linearly with \(w\) reaching \(f\drag \approx f_*\) at
\(w = \sound \approx \SI{10}{km.s^{-1}}\), then transitions to a quadratic
increase in the supersonic regime:
\begin{equation}
  \label{eq:fdrag-epstein}
  f\drag \approx
  \begin{cases}
    \text{if \(\phi^2 \ll 1\) and \(w_{10} < 1\):} & w_{10}\, f_* \\
    \text{if \(\phi^2 \ll 1\) and \(w_{10} > 1\):} & w_{10}^2\, f_* \\
  \end{cases}
  \ ,
\end{equation}
where \(w_{10} = w / \SI{10}{km.s^{-1}}\).

As \(\abs{\phi}\) increases, long-range electrostatic interactions with
protons within the Debye radius (Coulomb drag) become increasingly
important at subsonic relative velocities, as shown in
Figure~\ref{fig:gas-grain-drag-photoionized} by the orange
(\(\phi = 1\)), green (\(\phi = 4\)), and red (\(\phi = 16\)) lines.  However,
the Coulomb drag has a peak when \(w\) is equal to the thermal
speed of the colliders, which is \(\approx \SI{10}{km.s^{-1}}\) for protons,
giving a maximum strength of
\begin{equation}
  \label{eq:fdrag-maximum}
  f_{\mathrm{max}} = 0.5\, (\ln\Lambda)\, \phi^2 f_* \approx 10\, \phi^2 f_* \ , 
\end{equation}
where \(\Lambda\) is the plasma parameter (number of particles within a
Debye volume), such that
\(\ln\Lambda = 23.267 + 1.5 \ln T_4 - 0.5 \ln n\).  At highly super-thermal
speeds, the Coulomb drag falls asymptotically as
\(f\drag \propto 1 / w^{2}\).  The thermal speed of electrons is higher than
that of the protons by a factor of \((m_p / m_e)^{1/2}\), so that the
electron Coulomb drag gives a second peak of similar strength, but at
\(w \approx \SI{430}{km.s^{-1}}\).

@@

In Appendix~\ref{sec:gas-free-bow} we calculate the behaviour of an
incoming stream of dust grains, subject only to the repulsive
radiative force from a star.  We are now in a position to see how
gas--grain drag will modify this picture.  From
equations~\eqref{eq:dust-rad-force} and~\eqref{eq:fstar}, we can write
the radiation force acting on a grain as
\begin{equation}
  \label{eq:frad-Xi}
  f\rad = \Qp\, \Xi\, f_* \ ,
\end{equation}
where \(\Qp\) is the grain's radiation pressure efficiency (see
footnote~\ref{fn:Qp} in Appendix~\ref{sec:gas-free-bow}) and \(\Xi\) is
the local radiation parameter, defined as the ratio of direct stellar
radiation pressure to gas pressure:
\begin{equation}
  \label{eq:Xi-Prad-over-Pgas}
  \Xi \equiv \frac{P\rad}{P\gas} \approx \frac{L}{4 \pi R^2 c\, (2 n k T)} \ ,
\end{equation}
where the last expression corresponds to the optically thin limit and
pure ionized hydrogen.  The grain potential \(\phi\) is also primarily
determined by \(\Xi\), as shown in Appendix~\ref{sec:cloudy-models-dust}
and the inset graph of Figure~\ref{fig:gas-grain-drag-photoionized},
but with a slow dependence, which can be approximated\footnote{%
  There is a slight secondary dependence on the grain composition and
  stellar spectrum.  The relationship given in
  eq.~\eqref{eq:phi-vs-Xi} is appropriate for graphite grains and for
  stellar effective temperatures in the range \SIrange{20}{30}{kK}.
  For hotter stars than this, \(\phi\) should be multiplied by a further
  factor of \(1.5\), while for silicate grains it should be divided by
  \(1.5\).  } %
as
\begin{equation}
  \label{eq:phi-vs-Xi}
  \phi \approx 1.5 \bigl( 2.3 +  \ln \Xi \bigr) \ .
\end{equation}

>>>>

\begin{figure}
  \includegraphics[width=\linewidth]{figs/test-Fdrag-components}
  \caption{Contributions of different collider species to the
    dimensionless drag force, \(f\drag / f_*\), as a function of
    gas--grain slip velocity, \(w\).  Solid lines show the Coulomb
    (electrostatic) drag, while dashed lines show the Epstein
    (solid-body) drag.  Results are shown for dimensionless grain
    potential \(\phi = 10\).  All Coulomb forces scale with
    \(\phi^2\), while the Epstein forces are independent of \(\phi\).  The
    species labelled ``CNO++'' represents the combined effect of all
    metals (total abundance \num{8.5e-4}, effective atomic weight
    \num{15.3}), which are assumed to be doubly ionized.}
  \label{fig:drag-components}
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]{figs/test-Fdrag-param-space}
  \caption{Regimes of gas--grain drag as a function of slip velocity
    and grain potential.  The different regimes are indicated by bold
    roman numerals, as explained in
    Table~\ref{tab:fdrag-regimes}. Blue shading indicates regions
    dominated by Epstein (solid-body) drag, whereas red and green
    shading indicate regions dominated by Coulomb drag due to protons
    and electrons, respectively.  In each case, the saturated color
    represents a contribution \(> 70\%\) of the relevant component to
    the total drag force, while progressively lighter shading
    represents the \(> 60\%\) and \(> 50\%\) levels.  The thick white
    dotted line indicates the transition between the subthermal and
    superthermal regimes for protons, while the thin white dotted line
    indicates the corresponding transition for electrons.  Contours
    show the total drag force in units of \(f_*\) (see
    eq.~\eqref{eq:fstar}) in decade intervals from \(0.1\) to
    \(10^4\), as labelled.  Results are shown for
    \(T = \SI{8000}{K}\) and \(n = \SI{100}{cm^{-3}}\), but the
    differences are very slight throughout the ranges
    \(T = \text{\SIrange{5000}{15000}{K}}\) and
    \(n = \text{\SIrange{e-3}{e6}{cm^{-3}}}\).}
  \label{fig:drag-v-phi-plane}
\end{figure}

@@

\subsubsection{Drag force on grains}
\label{sec:drag-force-grains}

\begin{table}
  \centering
  \caption{Regimes of drag force as function of grain potential and slip speed}
  \label{tab:fdrag-regimes}
  \renewcommand\arraystretch{1.3}
  \resizebox{\linewidth}{!}{%
    \begin{tabular}{@{}r l l l@{}}
    \toprule
      & Regime & Approximate criteria & \(f\drag / f_*\) \\ \midrule
      I & Epstein subsonic & \(\phi^2 \ll 1\)
                             and \(w_{10} < 1\) & \(1.5\, w_{10}\) \\
      II & Epstein supersonic & \(w_{10} > 1\)
                                and \(w_{10} > 5\,\abs{\phi}\)& \( w_{10}^2\) \\
      III & Coulomb p\(^+\) subthermal & \(\phi^2 > 1\)
                                         and \(w_{10} < 1\) & \((1 + 20\, \phi^2)\,
                                                              w_{10}\) \\
    % & Coulomb p\(^+\) peak & \(\phi^2 > 1\)
    %                          and \(w_{10} \approx 1\) & \(1 + 10\, \phi^2\) \\
      IV & Coulomb p\(^+\) superthermal & \(\phi^2 > 1\)
                                          and \(1 < w_{10} < 5\) & \(w_{10}^2
                                                                   + 10\, \phi^2/w_{10}^2 \) \\
      V & Coulomb e\(^-\) subthermal & \(\phi^2 > 20\)
                                 and \(5 < w_{10} < 42\) & \(0.48\, \phi^2 \,
                                                           w_{10}\) \\
    \bottomrule
  \end{tabular}
  }
\end{table}

The drag force on a charged dust grain moving at a relative speed
\(w\) through a plasma has contributions from both direct collisions
and from electrostatic Coulomb interactions with ions and electrons.
We use the expressions in \citet{Draine:1979a}, equations~(4)--(6),
summing over protons, electrons, and helium ions.\footnote{Helium is
  assumed to be 50\% neutral and 50\% singly ionized, leading to only
  a small contribution to the drag force.  For much hotter stars, such
  as the central stars of planetary nebulae, helium may be doubly
  ionized, which leads to a more significant drag contribution,
  especially around \(w \approx \SI{5}{km.s^{-1}}\).} %
Results are shown in Figure~\ref{fig:gas-grain-drag-photoionized} for
different values of the dimensionless grain potential:
\(\phi = e^2 Z\grain / a kT\), which is the electrostatic potential
energy of a unit charge at the surface of a grain of charge
\(Z\grain\) and radius \(a\), in units of the characteristic thermal
energy of a gas particle.  The drag force \(f\drag\) is put in dimensionless units by dividing by
a characteristic force:
\begin{equation}
  \label{eq:fstar}
  f_* = 2 n k T \cdot \pi a^2 \ , 
\end{equation}
which is approximately\footnote{%
  To simplify the exposition, the gas pressure in this section is
  calculated assuming a fully ionized, pure hydrogen plasma, yielding
  \(P\gas = 2 n k T\). For typical ISM abundances, the contribution of
  helium and its corresponding electrons yield a correction to this of
  order 5\%.  The required modifications when a cool star interacts
  with a predominantly neutral gas stream are discussed later.  } %
the ionized gas pressure multiplied by the grain geometric cross
section.

For grains with low electric charge, \(\phi^2 \ll 1\), the drag force is
dominated by direct collisions of protons with the grain.  The gas
collisional mean free path is much larger than the grain size, so the
drag is in the Epstein regime \citep{Weidenschilling:1977b}.  This is
illustrated by the \(\phi = 0.25\) case (blue line) in
Figure~\ref{fig:gas-grain-drag-photoionized}. As the relative
gas--grain slip speed, \(w\), increases, \(f\drag\) first increases
linearly with \(w\) reaching \(f\drag \approx f_*\) at
\(w = \sound \approx \SI{10}{km.s^{-1}}\), then transitions to a quadratic
increase in the supersonic regime.

As \(\abs{\phi}\) increases, long-range electrostatic interactions with
protons within the Debye radius (Coulomb drag) become increasingly
important at subsonic relative velocities, as shown in
Figure~\ref{fig:gas-grain-drag-photoionized} by the orange
(\(\phi = 1\)), green (\(\phi = 4\)), and red (\(\phi = 16\)) lines.  However,
the Coulomb drag has a peak when \(w\) is equal to the thermal
speed of the colliders, which is \(\approx \SI{10}{km.s^{-1}}\) for protons,
giving a maximum strength of
\begin{equation}
  \label{eq:fdrag-maximum}
  f_{\mathrm{max}} = 0.5\, (\ln\Lambda)\, \phi^2 f_* \approx 10\, \phi^2 f_* \ , 
\end{equation}
where \(\Lambda\) is the plasma parameter (number of particles within a
Debye volume), such that
\(\ln\Lambda = 23.267 + 1.5 \ln T_4 - 0.5 \ln n\).  At highly super-thermal
speeds, the Coulomb drag falls asymptotically as
\(f\drag \propto 1 / w^{2}\).  The thermal speed of electrons is higher than
that of the protons by a factor of \((m_p / m_e)^{1/2}\), so that the
electron Coulomb drag gives a second peak fof similar strength, but at
\(w \approx \SI{430}{km.s^{-1}}\).  The behavior of \(f\drag\) in all these
different regimes is summarised in Table~\ref{tab:fdrag-regimes}, in
terms of \(\phi\) and \(w_{10} = w / \SI{10}{km.s^{-1}}\).

% \begin{equation}
%   \label{eq:fdrag-regimes}
%   f\drag \approx
%   \begin{cases}
%     \text{Epstein subsonic (\(\phi^2 \ll 1\) and \(w_{10} < 1\)):}
%     & w_{10}\, f_* \\
%     \text{Epstein supersonic (\(\phi^2 \ll 1\) and \(w_{10} > 1\)):}
%     & w_{10}^2\, f_* \\
%     \text{Coulomb p\(^+\) subthermal (\(\phi^2 > 1\) and \(w_{10} < 1\)):}
%     & (1 + 20 \phi^2)\, w_{10}\, f_* \\
%     \text{Coulomb p\(^+\) peak (\(\phi^2 > 1\) and \(w_{10} \approx 1\)):}
%     & (1 + 10 \phi^2)\, f_* \\
%     \text{Coulomb p\(^+\) superthermal (\(\phi^2 > 1\) and \(1 < w_{10} < 5\)):}
%     & (w_{10}^2 + 10 \phi^2/w_{10}^2) \, f_* \\
%     \text{Coulomb e\(^-\) subthermal (\(\phi^2 > 20\) and \(5 < w_{10} < 42\)):}
%     & 0.48 \phi^2 \, w_{10}\, f_* \\
%   \end{cases}
% \end{equation}

@@

\subsubsection{Gas--grain separation: drift and rip}
\label{sec:gas-grain-separ}

In Appendix~\ref{sec:gas-free-bow} we calculate the behaviour of an
incoming stream of dust grains, subject only to the repulsive
radiative force from a star.  We are now in a position to see how
gas--grain drag will modify this picture.  From
equations~\eqref{eq:dust-rad-force} and~\eqref{eq:fstar}, we can write
the radiation force acting on a grain as
\begin{equation}
  \label{eq:frad-Xi}
  f\rad = \Qp\, \Xi\, f_* \ ,
\end{equation}
where \(\Qp\) is the grain's radiation pressure efficiency (see
footnote~\ref{fn:Qp} in Appendix~\ref{sec:gas-free-bow}) and \(\Xi\) is
the local radiation parameter, defined as the ratio of direct stellar
radiation pressure to gas pressure:
\begin{equation}
  \label{eq:Xi-Prad-over-Pgas}
  \Xi \equiv \frac{P\rad}{P\gas} \approx \frac{L}{4 \pi R^2 c\, (2 n k T)} \ ,
\end{equation}
where the last expression corresponds to the optically thin limit.
The grain potential \(\phi\) is also primarily determined by \(\Xi\), as
shown in Appendix~\ref{sec:cloudy-models-dust} and the inset graph of
Figure~\ref{fig:gas-grain-drag-photoionized}, but with a slow
dependence, which can be approximated\footnote{%
  There is a slight secondary dependence on the grain composition and
  stellar spectrum.  The relationship given in
  eq.~\eqref{eq:phi-vs-Xi} is appropriate for graphite grains and for
  stellar effective temperatures in the range \SIrange{20}{30}{kK}.
  For hotter stars than this, \(\phi\) should be multiplied by a further
  factor of \(1.5\), while for silicate grains it should be divided by
  \(1.5\).  } %
as
\begin{equation}
  \label{eq:phi-vs-Xi}
  \phi \approx 1.5 \bigl( 2.3 +  \ln \Xi \bigr) \ .
\end{equation}

In the outer regions of the photoionized volume around an OB star,
close to the ionization front, the radiation parameter is low, with
typical value \(\Xi \sim 0.1\).  In this regime, the negative charge
current at the grain surface due to electron collisions is roughly in
balance with the positive current due to the ultraviolet photoelectric
effect \citep{Weingartner:2001b}, leading to a low grain potential,
\(\abs{\phi} < 1\), which may be positive or negative.  The low
\(\Xi\) means that the radiative force is also weak:
\(f\rad \sim 0.1 f_*\) from equation~\eqref{eq:frad-Xi} if
\(\Qp \sim 1\) at UV wavelengths, which is true for all but the smallest
grains.  Thus, from equation~\eqref{eq:fdrag-epstein}, the radiative
force can be balanced by Epstein drag if \(w_{10} \sim 0.1\), leading to
a small drift velocity, \(w\drift \sim \SI{1}{km.s^{-1}}\), of the grains
with respect to the gas.

As a dust grain
approaches a star, the radiation parameter \(\Xi\) will increase as
\(R^{-2}\)
